---
title: "Preliminary_model"
format: html
editor: visual
---

# Loading data

```{r}
# Loading Football Data Set
library(nflreadr)
library(tidyverse)
football_games <- load_schedules(2020:2023)
# Filter the dataset for games played in SoFi
la_games <- football_games |> filter(stadium == "SoFi Stadium")
game_days <- la_games |>
  distinct(gameday) |>
  mutate(game_day = 1)
# Convert to Date type
game_days$gameday <- as.Date(game_days$gameday)
```

```{r}
# Loading DV Data Set
la_data <- read_csv("Domestic_Violence_Calls_from_2020_to_Present_20250415.csv")
la_data <- janitor::clean_names(la_data)
domestic_violence <- la_data |> filter(grepl("INTIMATE", crm_cd_desc, ignore.case = TRUE))
# Convert to Date by specifying format and removing time
domestic_violence$date_occ <- 
  as.Date(domestic_violence$date_occ, format = "%m/%d/%Y %I:%M:%S %p")
```

```{r}
# Joining data sets
dv_with_games <- domestic_violence |>
  left_join(game_days, by = c("date_occ" = "gameday")) |>
  mutate(game_day = ifelse(is.na(game_day), 0, game_day))
# Creating data set of count of dv reports, grouped by date
dv_daily <- domestic_violence |>
  group_by(date_occ) |>
  summarize(dv_reports = n())
# Joining data set into our original big data set dv_with_games
dv_with_games <- dv_with_games |>
  group_by(date_occ) |>
  mutate(dv_reports = n())
```

# Preliminary Model

### Research Question

Do NFL games at SoFi Stadium have a measurable effect on the frequency of reported domestic violence incidents in Los Angeles? The project is trying to explore whether the is a statistically significant and potentially predictable relationship between spikes in domestic violence cases and NFL game days.

```{r}
# Joined Data Set
head(dv_with_games)
```

For our final project I plan to do these 3 methods, with possibly clustering if time permits.

```{r}
# Regression

# Cross-Validation

# Model Selection
```
